---
title: "Time-varying Mediation Function for Continuous Outcome 3 Treatment (Exposure Groups)"
author: "Yajnaseni Chakraborti"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Formatting Data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tvmediation)
library(locpol)
library(dplyr)
library(ggplot2)
```

<div style="text-align: justify">

The purpose of this vignette is to provide users with a step-by-step guide for 
performing and interpreting the results of a time varying mediation analysis for continuous outcome with 3 treatment (exposure) groups. We will rely on a dataset simulated (based on the Wisconsin Smokers' Health Study 2 dataset) which includes 1086 individuals across 3 treatment conditions. In one condition, one-third of participants received only a nicotine patch. In a second condition, another one-third received a nicotine patch and varenicline. A final third of participants were placed into a third condition and received a nicotine patch as well as combination nicotine replacement therapy. The  outcome variables of interest was 'Cessation Fatigue' that recorded . In addition, mediator variables were collected by asking participants if they felt a negative mood in the last fifteen minutes, and whether they wanted to smoke in the last 15 minutes. Both the 
outcome and mediator variables were assessed 2x per day over the course of 
twenty-eight days.
A traditional approach to analyzing this type of data would utilize a mediation analysis. First, a single direct effect would be calculated by regressing the outcome on the treatment condition. Next, a single indirect effect would be computed by multiplying the effect of treatment condition on the mediator by the effect of the mediator on the outcome. However, we feel this method potentially misses important information about the dynamic effect that a mediator may have over time. Specifically, we hypothesize that mood changes and thus, its mediating effect on one’s feelings of quitting smoking is likely to vary. We therefore propose a time varying mediation analysis which estimates the 
mediation effect as a function that varies over time.

## Getting started

To use the time varying mediation analysis package in R, you must first install
it and load it: 
```{r, eval = FALSE}
devtools::install.packages("tvmediation")
library(tvmediation)
library(locpol)
library(dplyr)
library(ggplot2)
```

The `tvma_3trt` function requires 4 necessary and 4 optional input variables.

1. `NRT1` A binary vector with treatment1 schedule
2. `NRT2` A binary vectot with treatment2 schedule
3. `t.seq` The time sequence of the measures
4. `mediator` The matrix of mediator values in wide format
5. `outcome` The matrix of outcome outcomes in wide format

The optional inputs are:

6. `t.est` The time sequence of estimation
7. `plot` TRUE or FALSE for plotting mediation effect. The default value is "FALSE".
8. `CI` "none" or "boot" method of deriving confidence intervals. The default value is "boot".
9. `replicates` Number of replicates for bootstrapping confidence intervals. The default value is 500.
10. `verbose` TRUE or FALSE for printing results to screen. The default value is "FALSE".

## Formatting your data before calling the `tvma_3trt` function

When there are three treatment groups, create two dummy variables `NRT1` and `NRT2` corresponding to the treatment (exposure) groups other than the placebo or the standard of care in this case. In our data, `NRT1 = 1` indicates that the participant has been given `varenicline` and `NRT1 = 0` for `no varenicline`. Similarly, `NRT2 = 1` means particpant has recieved the `combination treatment` and `NRT2 = 0` means particpant did not recieve the `combination treatment (comboNRT)`. Naturally, `NRT1 = 1` and `NRT2 = 1` are mutually exclusive that is a particpant cannot have `NRT1 = 1` and `NRT2 = 1` simultaneously. If `NRT1 = 1` then for that particpant `NRT2` must be equal to `0`. However, `NRT1 = 0` and `NRT2 = 0` are not mutually exclusive. When `NRT1 = 0` and `NRT2 = 0` then it is obvious that the particpant has recieved the placebo or the standard of care, in this case `patch`. Thus if there are three treatment groups, two dummy variables clearly indicating `NRT1` or `treatment1`, `NRT2` or `treatment2` and `placebo` need to be created. An example is given as follows. In the `smoker.rda` dataset, three columns indicating the assignment of the three treatments are present.

To load the simulated dataset `smoker.rda`, type:
```{r}
data(smoker)
```

Creating two dummy variables to indicate whether a particpant was given varenicline or comboNRT. `NRT1` indicates whether a subject recieved `Varenicline` or not. `NRT2` indicates whether a subject recieved `comboNRT` or not. Each subject's response for the treatment group (e.g. `varenicline`) was converted to a numeric value, and 1 was subtracted to yield a vector of zeros and ones.

```{r}
NRT1 <- as.numeric(unique(smoker[ ,c("SubjectID","varenicline")])[,2])-1
NRT2 <- as.numeric(unique(smoker[ ,c("SubjectID","comboNRT")])[,2])-1
```

Unlike the `tvmb` function for binary outcome or the `tvma` function for continuous outcome and 2 treatment groups, the user does not need to compare the mediation effects in parts, i.e. `varenicline` vs `placebo (nicotine patch)`, or `comboNRT` vs `placebo`, or `varenicline` vs `comboNRT`, a final variable was derived based on these two variables. The mathematical model is built to support time varying mediation for three exposure groups. For more information on the mathematical model, please read Cai et.al. 2020 (site the paper properly once published).

The `smoker` dataset is organized in Long format with `SubjectID` repeating
over multiple rows for each participant. The `tvmb` function requires that our
data be in Wide format in order to accurately estimate the time varying mediator
coefficients. The `tvmediation` package includes a useful function `LongToWide` 
to help users properly format their data for analysis.

`LongToWide` has 3 main input arguments, and a fourth optional argument.

1. `subject.id` requires a column of subject identifiers. 
2. `time.sequences` requires a column of time points.
3. `outcome` requires a column of measures that are to be transposed.
4. `verbose` is an option that can be turned on to print the output of `LongToWide` to the console

The output of `LongToWide` is a matrix of data in wide format where columns 
represent time sequence and rows represent the outcome for each subject.

As the `tvma_3trt` function will require two matrices, one for mediators, and one for
outcomes, we will use the `LongToWide` function twice as seen below:

```{r}
mediator <- LongToWide(smoker$SubjectID, smoker$timeseq, smoker$NegMoodLst15min)
outcome <- LongToWide(smoker$SubjectID, smoker$timeseq, smoker$cessFatig)
```

The `tvma_3trt` also requires the time sequence variable in a vector format.
```{r}
t.seq <- sort(unique(smoker$timeseq))
```

We are now ready to perform our time varying mediation analysis.
```{r, fig.width = 7, fig.height = 5, results='hide', fig.keep='all'} 
results_tvma_3trt <- tvma_3trt(NRT1, NRT2, t.seq, mediator, outcome, plot = TRUE)
```
```{r, echo = FALSE}
results_tvma_3trt
```

## Results

The `tvma_3trt` function returns a list of results including:

1. `hat.alpha1` the estimated effect of treatment1 on the mediator
2. `hat.alpha2` the estimated effect of treatment2 on the mediator
3. `hat.beta3` the estimated effect of the mediator on the outcome
4. `hat.mediation1` the time varying mediation effect of treatment1
5. `hat.mediation2` the time varying mediation effect of treatment2
6. `CI.lower.alpha1` the lower confidence intervals for hat.alpha1
7. `CI.upper.alpha1` the upper confidence intervals for hat.alpha1
8. `CI.lower.alpha2` the lower confidence intervals for hat.alpha2
9. `CI.upper.alpha2` the upper confidence intervals for hat.alpha2
10. `CI.lower.beta3` the lower confidence intervals for hat.beta3
11. `CI.upper.beta3` the upper confidence intervals for hat.beta3

Optional returns based on arguments passed by user include:

12. `CI.lower.NRT1` the lower confidence intervals for treatment1
13. `CI.upper.NRT1` the upper confidence intervals for treatment1
14. `CI.lower.NRT2` the lower confidence intervals for treatment2
15. `CI.upper.NRT2` the upper confidence intervals for treatment2
16. `SE_MedEff1` estimated standard error of mediation effect for treatment1
17. `SE_MedEff2` estimated standard error of mediation effect for treatment2

Returns if `plot = TRUE` argument is passed

18. `plot1_a1` plot for hat.alpha1 across t.seq
19. `plot2_a2` plot for hat.alpha2 across t.seq
20. `plot3_b3` plot for hat.beta3 across t.seq
21. `MedEff_NRT1` plot for medEffect across t.seq for treatment1
22. `MedEff_NRT2` plot for medEffect across t.seq for treatment2
23. `MedEff_CI_NRT1` plot for CIs of medEffect across t.seq for treatment1
24. `MedEff_CI_NRT2` plot for CIs of medEffect across t.seq for treatment2

As noted above, the CIs are only returned when bootstrap confidence intervals are computed; the plot results are only returned when `plot = TRUE` was passed as an argument. 

We recommend using the plot to interpret your findings as it may be difficult to derive meaning from the numerical values alone. Further, plots with confidence intervals and mediation effect from bootstrapped samples, are returned if `CI = "boot"` is passed.

For example, specifying `t.est`, and `replicates` in the follow code will produce 
estimates as time points 0.2, 0.4, 0.6, and 0.8 using 500 bootsrap replicates.
```{r, fig.width = 6, fig.height = 4, results = 'hide', fig.keep = 'all'}
results <- tvma_3trt(NRT1, NRT2, t.seq, mediator, outcome, t.est = c(0.2, 0.4, 0.6, 0.8), replicates = 500)
```
```{r, echo = FALSE}
results
```

## Summary
TBD

</div>
---